
```{r setoptions}
opts_chunk$set(echo=TRUE)
##loading the necessary R packages
library(plyr)
library(ggplot2)
library(xtable)
```
Analyses of Daily Walking Activity of an Anonymous Individual
========================================================
### Study Summary

- Use of small portable devices to collect data on personal activites
- An individual's daily activity was monitored by counting the number of steps per every 5-minute interval per day

### Goals of the Study

1.  To manipulate this dataset to extract, summarize and present the statiscal summaries that are appropriate to our study questions.
2.  To integrate the manipulation processes and the resulting summaries in an R markdown file to generate a document that complies lierate statistcal programming.   


### 1.  Mean Total Number of Steps per Day

- Import the data and name it "walk" 
- Add a new variable "weekday" (weekdays or weekend) based on the date. 
 
**Generation of histogram and summary table**

- **histogram** showing the distribution of total daily steps
- **table** summarizing the daily walking activities

```{r histogram, fig.height = 4}
#read the data frame "activity.csv" in the working directory
walk <- read.csv("activity.csv")

#add new varibale "weekday" and append
weekday <- weekdays(as.Date(walk$date, format="%m/%d/%Y"))
walk <- cbind(walk, weekday)
walk$date <- as.character(walk$date)
walk$weekday <- as.character(walk$weekday)

sortweekdays <- function(x){
     x <- replace(x, !(x %in% c("Saturday", "Sunday")), "Weekdays")
     x <- replace(x, x %in% c("Saturday", "Sunday"), "Weekend")
     }
walk$weekday <- sortweekdays(walk$weekday)

summaryTable  <- ddply(walk, c("date"), summarize,
     total=sum(steps, na.rm=T),
     average=round(sum(steps, na.rm=T)/length(steps)),
     median=median(steps, na.rm=T))
     
g <- ggplot(summaryTable, aes(total)) + 
     geom_histogram(binwidth=2000, fill="steelblue", col="red")
g <- g + ggtitle("Histogram of Total Steps per Day") +
  xlab("Total Steps") + ylab("Frequency")
print(g)
```

```{r summaryTable}
print(summaryTable, type="html")
```

### 2.  Pattern of Average Walkign Activity 

- Does this individual's walking activity show any kind of pattern?  
1.  Split the "activity (walk)" dataset by the "interval" variable
2.  Calculate the average steps for each 5-minute intervals

- **Time-series plot** showing the average steps per intervals 

```{r timeseries, fig.height=4}
## new data frame "stepPerInterval" splitted by "interval"
stepsPerInterval <- ddply(walk, .(interval), summarize, 
     average=mean(steps, na.rm=T))
g <- ggplot(stepsPerInterval, aes(interval, average)) + geom_line()
g <- g + ggtitle("Avergae Steps per Interval") +
     xlab("Interval") + ylab("Average Steps")

maxStep <- max(stepsPerInterval$average, na.rm=T)
maxInterval <- 
     stepsPerInterval$interval[stepsPerInterval$average==maxStep]
g <- g + geom_text(aes(maxInterval, maxStep), label=maxInterval, 
      color="red")
print(g)
```

#### This time series plot shows that this invidual's maximum walking activity is at the **`r maxInterval`th interval**.  

### 3.  Missing Data Imputation 
```{r naCount, fig.height=4}
naCount <- sum(is.na(walk$steps))
```
- The "activity" data set has total **`r naCount`** missing values in the "steps" variable.  
- The missing values were discrded in the above analyses.  
- This process resulted in a very unlikely outcome 
(For example 8 days with zero ealking activity)

**Imputation method**
- Split by "weekday"
- Split by "interval"
- Calculate the means per interval
- Replace NAs with the means of the corresponding intervals 

**I used ddply() function of the *plyr* package to split the data.  The replacena() function is introduced to replace na's**

```{r imputation, fig.height=4}
replacena <- function(x) {
      replace(x, is.na(x), round(mean(x, na.rm=T)))
      }

imputedwalk <- ddply(walk, c("weekday", "interval"), transform, 
     steps=replacena(steps))
imputedwalk <- imputedwalk[order(as.Date(imputedwalk$date), 
     imputedwalk$interval), ]

summaryTable1  <- ddply(imputedwalk, c("date"), summarize,
     total=sum(steps, na.rm=T),
     average=round(mean(steps, na.rm=T)),
     median=median(steps, na.rm=T))
```
- The resulting **histogram** and **Summary Table** are shown below.  
- The number of days with < 2,000 steps went down from 10 to 2 days   
- A noticeable change in the summary rable is in the median value.

```{r histogram2, fig.height=4} 
h <- ggplot(summaryTable1, aes(total)) + 
     geom_histogram(binwidth=2000, fill="steelblue", col="red")
h <- h + ggtitle("Histogram of Total Steps per Day (Imputed)") +
  xlab("Total Steps") + ylab("Frequency")
print(h)
```
```{r summaryTable1}
print(summaryTable1, type="html")
```

### 4.  Comparison of Activity Patterns between Weekdays and Weekends

```{r timeseries2, fig.height=6}
## new data frame stepPerInterval2 splitted by "weekday" & "interval"
stepsPerInterval2 <- ddply(imputedwalk, .(weekday, interval),
     summarize, average=mean(steps, na.rm=T))
f <- ggplot(stepsPerInterval2, aes(interval, average)) + geom_line()
f <- f + facet_grid(weekday~.)
f <- f + ggtitle("Avergae Steps per Interval (imputed)") +
     xlab("Interval") + ylab("Average Steps")
print(f)
```